Convergent Evolution During Local Adaptation to Patchy Landscapes

### 2014-12-18 patchy-review-responses.tex

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We thank the editor and the reviewers for their helpful comments. We
have substantially revised our paper in response to these comments,
and think that it is substantially improved as a consequence. Below we
have a detailed response point-by-point to the reviewers comments. We
first address the consensus points highlighted by the editor.

@@

  \item Several reviewers pointed out that in approximating $\E[1-(1-p_e)^K] \approx \E[K] p_e$
    we were neglecting the case that $s_m \ll s_p$.
    This omission was sloppy on our part,
    which we have rectified; see the new equation~\eqref{eqn:migrate}.

  \item We have also improved our discussion of the limits to the theory,
    mostly in the (short) section \nameref{ss:assumptions}.
    The main point of note was that, indeed as Reviewer 3 pointed out,
    we were originally running simulations in regions where $s_m$ was too small for the theory to apply,
    resulting in $\migrate$ that was not monotonic with $s_m$.
    Happily, the natural restriction of $R \sqrt{s_m}/\sigma > 1$ coincides with the condition
    for $\migrate$ to decrease with $s_m$;
    we don't have good intuition for why the formula should increase in the region where it is inapplicable,
    but this seems beside the point.

\end{itemize}

In addition to the format required by PLoS, we have again provided a version of the manuscript with all the figures in the right places.  
Also note that in both pdfs, references and sections names are hyperlinks to the correct locations in the manuscript.

### 2014-12-18 patchy-selection-paper.tex
resubmitted

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{color}
\usepackage[hidelinks]{hyperref}

@@

%%%%%
% Responses to reviews:
\input{review-response-commands}
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\ifreviewresponses\else\renewcommand{\linelabel}[1]{}\fi
\iflatexml\reviewresponsesfalse\fi  % they don't work so well in html
% \newcommand{\responsefile}{pbio-reviews-19sept12-responses.tex}  % name of the review reponses file
\newcommand{\responsefile}{patchy-review-responses.tex}  % name of the review reponses file

@@

\ifsubmissionversion
    \usepackage{endfloat}
    \usepackage{cite}
    % Use the PLoS provided bibtex style
    \bibliographystyle{plos2009}
    \newcommand{\citep}[1]{\cite{#1}}
    \newcommand{\citet}[1]{\cite{#1}}
\else
    \usepackage{natbib}
    \bibliographystyle{abbrvnat}
\fi

@@

% \newcommand{\var}{\mathop{\mbox{Var}}}
% \newcommand{\Exp}{\mathop{\mbox{Exp}}}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\Exp}{Exp}
\DeclareMathOperator{\sn}{sn}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\one}{\mathbf{1}}
\newcommand{\calE}{\mathcal{E}}
\newcommand{\calC}{\mathcal{C}}
\newcommand{\dconv}{\xrightarrow{d}}
\newcommand{\deq}{\stackrel{\scriptscriptstyle{d}}{=}}
\newcommand{\given}{\,\vert\,}

@@

Also, to avoid introducing more parameters,
we assume population density $\rho$ is equal across the range (even in the ``poor'' patches).
If this is not the case, in most equations below, the population density that should be used is the density
on the poor patches; see \citet{lenormand2002limits} for a more thorough approach.
We further assume that the variance in offspring number is $\xi^2$, 
and that the mean squared distance between parent and child is $\sigma^2$
(i.e.\ $\sigma$ is the dispersal distance). 
We will deal with migration by immediately appealing to the central limit theorem,
treating the total distance traveled across $t$  dispersal events as Gaussian with variance $t \sigma^2$,
and do not discuss the (interesting) cases where rare, long-distance dispersal events
are more important \citep[see e.g.][for discussion]{levin2003ecology,ralph2010parallel,hallatschek2014acceleration}.  \linelabel{rr:longtailed}
%We also aim to describe the scenario where it is sensible to think of large, discrete patches;
%the case of a fine-scale heterogeneous environment is treated
%\citep{elsewhere}. \gc{IS THIS LAST SENTENCE NEEDED IF SO GIVE CITATION} \plr{haven't found it yet}

@@

\paragraph{Decay of shared haplotypes} \linelabel{rr:haplotype_discussion}
To provide context for the results on shared haplotype length in
section \secref{ss:haplotype_length}
it is important to also understand
the process by which haplotypes are whittled down within patches.
The initial haplotype that sweeps within a patch 
will be dispersed over time by recombination.
Likewise, the haplotype that is shared between patches coadapted by migration
(equation~\ref{eqn:haplotype_length}) will also break down. 
However, a long time after the initial sweep,
we may still expect to find individuals within the patch sharing longer haplotypes around the selected locus
than with individuals elsewhere,
since selection against migrants 
decreases mean coalescence times within the patch near the selected locus.  %\citep{bartonpaper,charlesworths}. 
The literature on clines \citep[e.g.][]{barton1979geneflow} has important information,
but more work is needed
to provide robust estimates for these processes.
Questions about the genomic length-scale of signals of
sweeps shared by migration have also been addressed in discrete
population settings \citep{slatkin1998hitchhiking,kim2011hitchhiking}, reviewed in \citet{barton2000genetic}. 
This work has shown that the length of the shared swept haplotype is often significantly
shorter than the sweep with each patch, resulting in a pattern of
shoulders of elevated $F_{ST}$ between adapted populations some distance away from
the shared selected allele. It would be of interest to see how similar
patterns can arise in a continuous population setting, as a way of
uniting these results.

@@

Here we describe how to solve \eqref{eqn:radial_eqfreq_diffeq} exactly in one dimension,
i.e.\ the Fisher-KPP equation with piecewise constant selection.
Let $\xi(t,x)$ be the proportion of the allele under spatially varying selection $s(x)$ in one dimension,
so that 
\[
\partial_t p(t,x) = \frac{\mu}{2} \sigma^2 \partial_x^2 p(t,x) + \mu s(x) p(t,x) (1-p(t,x)) .
\]
The stable distribution $\phi(x) = \lim_{t\to\infty} p(t,x)$ then solves
\begin{align} \label{eqn:definingphi}
    \partial_x^2 \phi(x) = - 2 s(x) \phi(x) (1-\phi(x)) /\sigma^2,
\end{align}
with appropriate boundary conditions.
First rescale space by $\sigma/\sqrt{2}$ so the $\sigma^2/2$ term disappears.
If we now assume that $s(x)$ is piecewise constant,
$s(x) = s_i$ for $x \in [x_i,x_{i+1})$, with $x_0=-\infty$ and $x_{n+2}=\infty$,
then the equation is integrable: if we multiply through by $2\partial_x \phi(x)$ and integrate, then we get that
\begin{align} \label{eqn:conservation}
    ( \partial_x \phi(x) )^2  &= - \int^{x} 2 s(x) \phi(x) (1-\phi(x)) \partial_x \phi(x) dx \\
        &= - s_i \phi(x)^2 \left( 1 - \frac{2}{3} \phi(x) \right) - K_i \quad \mbox{for } x \in [x_i,x_{i+1}) ,
\end{align}
if we define
\begin{align} 
  K_i = - s_i \phi(x_i)^2 \left( 1 - \frac{2}{3} \phi(x_i) \right) - ( \partial_x \phi(x_i) )^2  .
\end{align}
For ease of reference, we define
\[
        V_i(\phi) = s_i \phi^2 \left( 1 - \frac{2}{3} \phi \right) + K_i .
\]
Note that $V_i'(0)=V_i'(1)=0$, that $V_i(0)=K_i$ and $V_i(1) = K_i+s_i/3$.  
We will always have that $V(\phi) \le 0$.
(We have then that $\partial_x^2 \phi = - \partial_\phi V(\phi)$, the equation of motion of a particle in potential $V$.
Also note that this implies ``conservation of energy'', i.e.\ $( \partial_x \phi(x) )^2 + V(\phi(x))$ is constant.)
Rearranging, we get that $dx = d\phi / \sqrt{-V(\phi)}$, so where $\phi(x)$ is monotone, the inverse is
\[
    x(\phi) = x(\phi^*) \pm \int_{\phi^*}^\phi \frac{ d\psi }{ \sqrt{ -V(\psi) } } .
\]
For each $i$ then define the elliptic function
\begin{align}  \label{eqn:elliptic_function}
    F_i(\phi) = \int_{\phi_i^*}^\phi \frac{ d\psi }{ \sqrt{ -V_i(\psi) } } ,
\end{align}
where take the positive branch of the square root, and $\phi_i^*$ will be chosen later.
Then we have that
$x(\phi) - x(\phi_0) = \pm( F(\phi) - F(\phi_0))$,
or for an appropriate $x_0$,
\[
    \phi(x) = F^{-1}\left( F(\phi(x_0)) \pm (x - x_0) \right).
\]

@@

@@

% sm=0.1, R=80
\begin{figure}
  \begin{center}
    \includegraphics{example-migration-sims/89826-r1-501-sb0_01-sm-0_1-N4000-pophistory-run}
    \includegraphics{example-migration-sims/76178-r1-501-sb0_01-sm-0_1-N1000-pophistory-run}
    \includegraphics{example-migration-sims/97545-r1-501-sb0_01-sm-0_01-N100-pophistory-run}
  \end{center}
  \caption{
    Randomly chosen simulations of adaptation by migration
    with $s_m=0.1$, $R=80$, $\sigma\approx 1$, and $\rho$ varying.
    On the left of each is a space-time heatmap of the local frequency of $B$ alleles;
    and on the right are twenty-five curves showing the frequencies of $B$ at evenly spaced time points
    (i.e.\ each line represents a vertical slice through the plot on the left);
    dotted black lines indicate the patches where $B$ is advantageous.
  } \label{sfig:sims_5}
\end{figure}

@@

\begin{figure}[ht!]
    \begin{center}
        % made in patchy-sim-writeup-plots.R
        \includegraphics{example-equilibrium}
    \end{center}
    \label{fig:example_equil}
    \caption{
    Snapshots of a simulation of allelic frequency dynamics around a patch where the allele is beneficial:
    the focal allele is benefical in the middle region and deleterious outside.
    The dynamics were simulated for 10000 generations across demes with 1000 individuals in each
    and random dispersal to nearby demes with a mean dispersal distance of $\sigma \approx 2$ demes.
    Colored lines show allele frequencies across space in 50 uniformly spaced generations;
    the solid line shows the mean across all generations.
    Below is the same plot, except on a log scale.
    }
\end{figure}

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{color}
\usepackage[hidelinks]{hyperref}
\usepackage{etoolbox}

@@

%%%%%
% Responses to reviews:
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\iflatexml\reviewresponsesfalse\fi  % they don't work so well in html
% \newcommand{\responsefile}{pbio-reviews-19sept12-responses.tex}  % name of the review reponses file
\newcommand{\responsefile}{patchy-review-responses.tex}  % name of the review reponses file

% override above choices (see makefile)
\ifdef{\submissionnofigs}{  % for submission, no figures
  \reviewresponsesfalse
  \submissionversiontrue
}{
}
\ifdef{\submissionfigs}{  % for reviewers, with figs
  \reviewresponsestrue
  \submissionversionfalse
}{
}
\ifdef{\arxiv}{  % for general use
  \reviewresponsesfalse
  \submissionversionfalse
}{
}

\ifreviewresponses
  \input{review-response-commands}
\else
  \newcommand{\linelabel}[1]{}
  \newcommand{\revpoint}[2]{}
  \newcommand{\includereviews}{}
\fi

@@

\ifsubmissionversion
    \usepackage[tablesfirst]{endfloat}
    \usepackage{cite}
    % Use the PLoS provided bibtex style
    \bibliographystyle{plos2009}
    \newcommand{\citep}[1]{\cite{#1}}
    \newcommand{\citet}[1]{\cite{#1}}
\else
    \usepackage{natbib}
    \bibliographystyle{abbrvnat}
\fi

@@

\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\Exp}{Exp}
\DeclareMathOperator{\sn}{sn}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\one}{\mathbf{1}}
\newcommand{\calE}{\mathcal{E}}
\newcommand{\calC}{\mathcal{C}}
\newcommand{\dconv}{\xrightarrow{d}}
\newcommand{\deq}{\stackrel{\scriptscriptstyle{d}}{=}}
\newcommand{\given}{\,\vert\,}

@@

Also, to avoid introducing more parameters,
we assume population density $\rho$ is equal across the range (even in the ``poor'' patches).
If this is not the case, in most equations below, the population density that should be used is the density
on the poor patches; see \citet{lenormand2002limits} for a more thorough approach.
We further assume that the variance in offspring number is $\xi^2$, 
and that the mean squared distance between parent and child is $\sigma^2$
(i.e.\ $\sigma$ is the dispersal distance). 
We will deal with migration by immediately appealing to the central limit theorem,
treating the total distance traveled across $t$  dispersal events as Gaussian with variance $t \sigma^2$,
and do not discuss the (interesting) cases where rare, long-distance dispersal events
are more important (see e.g.\ \citet{levin2003ecology,ralph2010parallel,hallatschek2014acceleration} for discussion).  \linelabel{rr:longtailed}
%We also aim to describe the scenario where it is sensible to think of large, discrete patches;
%the case of a fine-scale heterogeneous environment is treated
%\citep{elsewhere}. \gc{IS THIS LAST SENTENCE NEEDED IF SO GIVE CITATION} \plr{haven't found it yet}

@@

\paragraph{Decay of shared haplotypes} \linelabel{rr:haplotype_discussion}
To provide context for the results on shared haplotype length in
section \secref{ss:haplotype_length}
it is important to also understand
the process by which haplotypes are whittled down within patches.
The initial haplotype that sweeps within a patch 
will be dispersed over time by recombination.
Likewise, the haplotype that is shared between patches coadapted by migration
(equation~\ref{eqn:haplotype_length}) will also break down. 
However, a long time after the initial sweep,
we may still expect to find individuals within the patch sharing longer haplotypes around the selected locus
than with individuals elsewhere,
since selection against migrants 
decreases mean coalescence times within the patch near the selected locus.  %\citep{bartonpaper,charlesworths}. 
The literature on clines (e.g.\ \citet{barton1979geneflow}) has important information,
but more work is needed
to provide robust estimates for these processes.
Questions about the genomic length-scale of signals of
sweeps shared by migration have also been addressed in discrete
population settings \citep{slatkin1998hitchhiking,kim2011hitchhiking}, reviewed in \citet{barton2000genetic}. 
This work has shown that the length of the shared swept haplotype is often significantly
shorter than the sweep with each patch, resulting in a pattern of
shoulders of elevated $F_{ST}$ between adapted populations some distance away from
the shared selected allele. It would be of interest to see how similar
patterns can arise in a continuous population setting, as a way of
uniting these results.

@@

Here we describe how to solve \eqref{eqn:radial_eqfreq_diffeq} exactly in one dimension,
i.e.\ the Fisher-KPP equation with piecewise constant selection.
Let $p(t,x)$ be the proportion of the allele under spatially varying selection $s(x)$ in one dimension,
so that 
\[
\partial_t p(t,x) = \frac{\mu}{2} \sigma^2 \partial_x^2 p(t,x) + \mu s(x) p(t,x) (1-p(t,x)) .
\]
The stable distribution $\phi(x) = \lim_{t\to\infty} p(t,x)$ then solves
\begin{align} \label{eqn:definingphi}
    \partial_x^2 \phi(x) = - 2 s(x) \phi(x) (1-\phi(x)) /\sigma^2,
\end{align}
with appropriate boundary conditions.
First rescale space by $\sigma/\sqrt{2}$ so the $\sigma^2/2$ term disappears.
If we now assume that $s(x)$ is piecewise constant,
$s(x) = s_i$ for $x \in [x_i,x_{i+1})$, with $x_0=-\infty$ and $x_{n+2}=\infty$,
then the equation is integrable: if we multiply through by $2\partial_x \phi(x)$ and integrate, then we get that
\begin{align} \label{eqn:conservation}
    ( \partial_x \phi(x) )^2  &= - \int^{x} 2 s(x) \phi(x) (1-\phi(x)) \partial_x \phi(x) dx \\
        &= - s_i \phi(x)^2 \left( 1 - \frac{2}{3} \phi(x) \right) - K_i \quad \mbox{for } x \in [x_i,x_{i+1}) ,
\end{align}
if we define
\begin{align} 
  K_i = - s_i \phi(x_i)^2 \left( 1 - \frac{2}{3} \phi(x_i) \right) - ( \partial_x \phi(x_i) )^2  .
\end{align}
For ease of reference, we define
\[
        V_i(\phi) = s_i \phi^2 \left( 1 - \frac{2}{3} \phi \right) + K_i .
\]
Note that $V_i'(0)=V_i'(1)=0$, that $V_i(0)=K_i$ and $V_i(1) = K_i+s_i/3$.  
We will always have that $V(\phi) \le 0$.
(We have then that $\partial_x^2 \phi = - \partial_\phi V(\phi)$, the equation of motion of a particle in potential $V$.
Also note that this implies ``conservation of energy'', i.e.\ $( \partial_x \phi(x) )^2 + V(\phi(x))$ is constant.)
Rearranging, we get that $dx = d\phi / \sqrt{-V(\phi)}$, so where $\phi(x)$ is monotone, the inverse is
\[
    x(\phi) = x(\phi^*) \pm \int_{\phi^*}^\phi \frac{ d\psi }{ \sqrt{ -V(\psi) } } .
\]
For each $i$ then define the elliptic function
\begin{align}  \label{eqn:elliptic_function}
    F_i(\phi) = \int_{\phi_i^*}^\phi \frac{ d\psi }{ \sqrt{ -V_i(\psi) } } ,
\end{align}
where take the positive branch of the square root, and $\phi_i^*$ will be chosen later.
Then we have that
$x(\phi) - x(\phi_0) = \pm( F(\phi) - F(\phi_0))$,
or for an appropriate $x_0$,
\[
    \phi(x) = F^{-1}\left( F(\phi(x_0)) \pm (x - x_0) \right).
\]

@@

\section*{Supplemental material}

% restart figure and table numbering
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}

@@

% sm=0.1, R=80
\begin{figure}
  \begin{center}
    \includegraphics{example-migration-sims/89826-r1-501-sb0_01-sm-0_1-N4000-pophistory-run}
    \includegraphics{example-migration-sims/76178-r1-501-sb0_01-sm-0_1-N1000-pophistory-run}
    \includegraphics{example-migration-sims/95095-r1-501-sb0_01-sm-0_1-N100-pophistory-run}
  \end{center}
  \caption{
    Randomly chosen simulations of adaptation by migration
    with $s_m=0.1$, $R=80$, $\sigma\approx 1$, and $\rho$ varying.
    On the left of each is a space-time heatmap of the local frequency of $B$ alleles;
    and on the right are twenty-five curves showing the frequencies of $B$ at evenly spaced time points
    (i.e.\ each line represents a vertical slice through the plot on the left);
    dotted black lines indicate the patches where $B$ is advantageous.
  } \label{sfig:sims_5}
\end{figure}

@@

\begin{figure}[ht!]
    \begin{center}
        % made in patchy-sim-writeup-plots.R
        % \includegraphics{example-equilibrium}
    \end{center}
    \label{fig:example_equil}
    \caption{
    Snapshots of a simulation of allelic frequency dynamics around a patch where the allele is beneficial:
    the focal allele is benefical in the middle region and deleterious outside.
    The dynamics were simulated for 10000 generations across demes with 1000 individuals in each
    and random dispersal to nearby demes with a mean dispersal distance of $\sigma \approx 2$ demes.
    Colored lines show allele frequencies across space in 50 uniformly spaced generations;
    the solid line shows the mean across all generations.
    Below is the same plot, except on a log scale.
    }
\end{figure}
